Statistical Considerations in Multilevel Mediation Analysis

### 2025-02-03 papers/Manuscript - Exact Asymptotics (William).tex
Copy manuscript for uploading to Overleaf
MacEwan Presentation

\maketitle
\section{Notes to Self}
\begin{outline}
    \1 In the paper, I organize variables as $Y, X, M, W$. This is based on defining counterfactuals as $Y_x$, then adding $M$, giving $Y_{x, m}$. In my code, it's always $Y, M, X, W$. I must be very careful to not mess this ordering up. The code is more arbitrary, but it's also harder to change. For now, I'm just going to try to juggle this difference manually, but I will need to carefully validate any results when passing from code to paper (this sort of issue has already been the source of one elusive error in my code).
    \1 I need to be very explicit that our analysis of expected counterfactuals is in service to mediation effects. Once we have estimates and UQ for expected counterfactuals, the extension to total, direct and indirect effects is straightforward.
\end{outline}

@@

It is a well-known fact about logistic regression that
%
\begin{equation}
    \bP \left( Y=1 |U, X=x, M=1, W=w \right) = \left[1 + \exp(-\eta_Y)\right]^{-1},
\end{equation}
%
so the first term in Equation (\ref{eq:ID_bin_ran}) can be written as
%
\begin{align}
    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_3(u; 0, \Gamma_Y)}{1 + \exp(-\eta_Y)} du, \label{eq:pY1_bad}
\end{align}
%
where $\phi_d$ is the $d$-variate normal density. A straightforward change of variables gives us the alternative expression
%
\begin{align}
    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_1(z; 0, 1)}{1 + \exp(-\mu_Y - \alpha_M - \gamma_Y(1, x, 1) z)} dz. \label{eq:pY1}
\end{align}
%
Importantly, the integral in Equation (\ref{eq:pY1}) is univariate (and thus amenable to numerical evaluation using quadrature). The form of this integral arises frequently enough that we give it a name. Let
%
\begin{equation}
    \Psi(a, b) = \int \frac{\phi(z; 0, 1)}{1 + \exp(-a - b z)} dz.
\end{equation}
%
We can now write the first term in \eqref{eq:ID_bin_ran} compactly as
%
\begin{equation}
    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) = \Psi(\mu_Y + \alpha_M, \gamma_Y(1, x, 1))
\end{equation}
%
Returning to the other terms in Equation \eqref{eq:ID_bin_ran}, similar expressions hold. Write $\eta_M = (\beta_0 + \beta_X x + B_W^T w) + (V_0 + V_X x)$ for the linear predictor of $M$ based on $X$ and $W$. Write $\mu_M = \beta_0 + \beta_X x + B_W^T w$ and $\xi_M = V_0 + V_X x$ for the fixed and random components respectively of $\eta_M$. Finally, write $\gamma^2_M(c_1, c_2) = (c_1, c_2) \Gamma_M (c_1, c_2)^T$, so that $\bV \xi_M = \gamma^2_M(1, x)$ ($\Gamma_M$ is the covariance matrix of $V$). We can now write the expected counterfactual in \eqref{eq:ID_bin_ran} as
%
\begin{align}
    \bE (Y_{x, M_{x'}} | W=w) = & \left[  \Psi(\mu_Y + \alpha_M, \gamma_Y(1, x, 1)) \cdot \Psi(\mu_M , \gamma_Y(1, x')) \right] + \label{eq:Psi_bin_ran}\\
    & \left[ \Psi(\mu_Y, \gamma_Y(1, x, 0)) \cdot \Psi(- \mu_M , \gamma_Y(1, x')) \right] \nonumber
\end{align}
%
Recall that each term in \eqref{eq:Psi_bin_ran} is a univariate integral, and can thus be accurately evaluated using standard numerical quadrature routines available in most software packages.

@@

Although not explicit in our notation, Equation \eqref{eq:Psi_bin_ran} depends only on the parameters of our two regression models (as well as the levels of our covariates, $x$, $x'$ and $w$). Estimation of the expected counterfactual thus proceeds by first fitting these two regression models to our data, then plugging-in our estimates of the two sets of model parameters to Equation \eqref{eq:Psi_bin_ran}. More precisely, if we re-write \eqref{eq:Psi_bin_ran} as
%
\begin{align}
    \bE (Y_{x, M_{x'}} | W=w) = \kappa(x, x', w; \alpha, \Gamma_Y, \beta, \Gamma_Y) \label{eq:kappa}\\
    &= \kappa(x, x', w), 
\end{align}
%
then our estimate of this expected counterfactual is
\begin{align}
    \hat{\bE} (Y_{x, M_{x'}} | W=w) &= \kappa(x, x', w; \hat{\alpha}, \hat{\Gamma}_Y, \hat{\beta}, \hat{\Gamma}_M) \label{eq:kappa_hat}\\
    &=: \hat{\kappa}(x, x', w)
\end{align}
%
We often suppress $w$ in both $\kappa$ and $\hat{\kappa}$.

Uncertainty quantification for our estimator, $\hat{\kappa}$, proceeds by the $\delta$-method \citep[see, e.g.,][]{vdV98}. We first get asymptotic covariance matrices for estimators of our two sets of regression parameters. Next, we stack these two covariances into a single $2\times 2$ block-diagonal matrix (\hl{assuming asymptotic independence between models, see} \citealp{Bau06}), then pre- and post-multiply by the Jacobian of $\kappa$. From traditional M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers (given appropriate regularity conditions). These objective functions differ from log-likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from its joint likelihood with the random-effects.

We omit here an explicit formula for the asymptotic covariance of $\hat{\kappa}$, \textcolor{red}{but one can be found in our \texttt{R} package\footnote{Name idea: \texttt{MultMedUQ}?}. See also the supplemental material for an accompanying derivation.} For concreteness however, write $\sqrt{n} (\hat{\kappa}(x, x') - \kappa(x, x')) \rightsquigarrow N(0, \lambda^2(x, x'))$. We will often also need the limiting joint dependence of $\kappa$ evaluated at two sets of values for $x$ and $x'$. In this case we write $\rho(x_1, x'_1; x_2, x'_2)$ for the asymptotic correlation between $\hat{\kappa}(x_1, x'_1)$ and $\hat{\kappa}(x_2, x'_2)$.

>>>>

\maketitle
\section{Notes to Self}
\begin{outline}
    \1 In the paper, I organize variables as $Y, X, M, W$. This is based on defining counterfactuals as $Y_x$, then adding $M$, giving $Y_{x, m}$. In my code, it's always $Y, M, X, W$. I must be very careful to not mess this ordering up. The code is more arbitrary, but it's also harder to change. For now, I'm just going to try to juggle this difference manually, but I will need to carefully validate any results when passing from code to paper (this sort of issue has already been the source of one elusive error in my code).
    \1 I need to be very explicit that our analysis of expected counterfactuals is in service to mediation effects. Once we have estimates and UQ for expected counterfactuals, the extension to total, direct and indirect effects is straightforward.
    \1 I have written ???? everywhere that I need to track-down a citation. This has the advantages of both showing up when I search for ??? (indicating a broken reference), and being different from ??? (so I can tell that it's deliberate, not a broken reference).
\end{outline}

@@

It is a well-known fact about logistic regression that
%
\begin{equation}
    \bP \left( Y=1 |U, X=x, M=1, W=w \right) = \left[1 + \exp(-\eta_Y)\right]^{-1},
\end{equation}
%
so the first term in Equation (\ref{eq:ID_bin_ran}) can be written as
%
\begin{align}
    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_3(u; 0, \Gamma_Y)}{1 + \exp(-\eta_Y)} du, \label{eq:pY1_bad}
\end{align}
%
where $\phi_d$ is the $d$-variate normal density. A straightforward change of variables gives us the alternative expression
%
\begin{align}
    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) & = \int \frac{\phi_1(z; 0, 1)}{1 + \exp(-\mu_Y - \alpha_M - \gamma_Y(1, x, 1) z)} dz. \label{eq:pY1}
\end{align}
%
Importantly, the integral in Equation (\ref{eq:pY1}) is univariate (and thus amenable to numerical evaluation using quadrature). The form of this integral arises frequently enough that we give it a name. Let
%
\begin{equation}
    \Psi(a, b) = \int \frac{\phi(z; 0, 1)}{1 + \exp(-a - b z)} dz.
\end{equation}
%
We can now write the first term in \eqref{eq:ID_bin_ran} compactly as
%
\begin{equation}
    \bE_U \bP \left( Y=1 |U, X=x, M=1, W=w \right) = \Psi(\mu_Y + \alpha_M, \gamma_Y(1, x, 1))
\end{equation}
%
Returning to the other terms in Equation \eqref{eq:ID_bin_ran}, similar expressions hold. Write $\eta_M = (\beta_0 + \beta_X x + B_W^T w) + (V_0 + V_X x)$ for the linear predictor of $M$ based on $X$ and $W$. Write $\mu_M = \beta_0 + \beta_X x + B_W^T w$ and $\xi_M = V_0 + V_X x$ for the fixed and random components respectively of $\eta_M$. Finally, write $\gamma^2_M(c_1, c_2) = (c_1, c_2) \Gamma_M (c_1, c_2)^T$, so that $\bV \xi_M = \gamma^2_M(1, x)$ ($\Gamma_M$ is the covariance matrix of $V$). We can now write the expected counterfactual in \eqref{eq:ID_bin_ran} as
%
\begin{align}
    \bE (Y_{x, M_{x'}} | W=w) = & \left[  \Psi(\mu_Y + \alpha_M, \gamma_Y(1, x, 1)) \cdot \Psi(\mu_M , \gamma_M(1, x')) \right] + \label{eq:Psi_bin_ran}\\
    & \left[ \Psi(\mu_Y, \gamma_Y(1, x, 0)) \cdot \Psi(- \mu_M , \gamma_M(1, x')) \right] \nonumber
\end{align}
%
Recall that each term in \eqref{eq:Psi_bin_ran} is a univariate integral, and can thus be accurately evaluated using standard numerical quadrature routines available in most software packages.

@@

Although not explicit in our notation, Equation \eqref{eq:Psi_bin_ran} depends only on the parameters of our two regression models (as well as the levels of our covariates, $x$, $x'$ and $w$). Estimation of the expected counterfactual thus proceeds by first fitting these two regression models to our data, then plugging-in our estimates of the two sets of model parameters to Equation \eqref{eq:Psi_bin_ran}. More precisely, if we re-write \eqref{eq:Psi_bin_ran} as
%
\begin{align}
    \bE (Y_{x, M_{x'}} | W=w) &= \kappa(x, x', w; \alpha, \Gamma_Y, \beta, \Gamma_Y) \label{eq:kappa}\\
    &= \kappa(x, x', w), 
\end{align}
%
then our estimate of this expected counterfactual is
\begin{align}
    \hat{\bE} (Y_{x, M_{x'}} | W=w) &= \kappa(x, x', w; \hat{\alpha}, \hat{\Gamma}_Y, \hat{\beta}, \hat{\Gamma}_M) \label{eq:kappa_hat}\\
    &=: \hat{\kappa}(x, x', w)
\end{align}
%
We often suppress $w$ in both $\kappa$ and $\hat{\kappa}$.

Uncertainty quantification for our estimator, $\hat{\kappa}$, proceeds by the $\delta$-method \citep[see, e.g.,][]{vdV98}. We first get asymptotic covariance matrices for estimators of our two sets of regression parameters. Next, we stack these two covariances into a single $2\times 2$ block-diagonal matrix (\hl{assuming asymptotic independence between models, see} \citealp{Bau06}), then pre- and post-multiply by the Jacobian of $\kappa$. From classical M-estimator theory, the asymptotic covariance of our regression estimators is approximately the inverse Hessian of their respective objective functions at the maximizers (given appropriate regularity conditions). These objective functions differ from log-likelihoods only in that Laplace's method is used to approximate the marginal likelihood of the data from its joint likelihood with the random-effects.

We omit here an explicit formula for the asymptotic covariance of $\hat{\kappa}$, \textcolor{red}{but one can be found in our \texttt{R} package\footnote{Name idea: \texttt{MultMedUQ}?}. See also the supplemental material for an accompanying derivation.} For concreteness however, write $\sqrt{n} (\hat{\kappa}(x, x') - \kappa(x, x')) \rightsquigarrow N(0, \lambda^2(x, x'))$, where $\rightsquigarrow$ denotes convergence in distribution. We will often also need the limiting joint dependence of $\kappa$ evaluated at two sets of values for $x$ and $x'$. In this case we write $\rho(x_1, x'_1; x_2, x'_2)$ for the asymptotic correlation between $\hat{\kappa}(x_1, x'_1)$ and $\hat{\kappa}(x_2, x'_2)$.

### 2025-02-03 presentations/MacEwan 2025-01/SSC 2024.tex
MacEwan Presentation

    Standard approach is multi-level modelling
    \begin{itemize}
        \item I.e. Add random effects which vary across clusters \newline
    \end{itemize}

@@

\begin{frame}{\GLMMs}
    The core idea is to augment our set of covariates
    \begin{itemize}
        \item Coefficients of these new covariates are random variables which vary across groups/clusters \newline
    \end{itemize}

    In the linear setting:
    \begin{itemize}
        \item Old model: $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \varepsilon$
        \item New model: $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \mathbf{u_1 Z_1 + \ldots + u_q Z_q} + \varepsilon$ \newline
    \end{itemize}

@@

    Strategies include:
    \begin{itemize}
        \item \textbf{Bootstrap}
        \item Quasi-Bayesian Monte Carlo
        \item $\delta$-method
    \end{itemize}
\end{frame}

>>>>

    Standard approach is multi-level modelling
    \begin{itemize}
        \item I.e. Add random effects that vary across clusters \newline
    \end{itemize}

@@

\begin{frame}{\GLMMs}
    The core idea is to augment our set of covariates
    \begin{itemize}
        \item Coefficients of these new covariates are random variables that vary across groups/clusters \newline
    \end{itemize}

    In the linear setting:
    \begin{itemize}
        \item Old model: \newline
        $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p + \varepsilon$
        \item New model:  \newline
        $Y = \alpha_0 + \alpha_1 X_1 + \ldots + \alpha_p X_p$ \newline
        \hspace{1cm} $\mathbf{+ u_1 Z_1 + \ldots + u_q Z_q} + \varepsilon$ 
    \end{itemize}

@@

    Strategies include:
    \begin{itemize}
        \item \textbf{Bootstrap} (Rado Ramasy, yesterday)
        \item Quasi-Bayesian Monte Carlo
        \item $\delta$-method
    \end{itemize}
\end{frame}
